OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-03-05 paper/paper.Rmd

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. 
**TODO: more words here?** 
Here, we tested the OptiFit algorithm with the reference as a database or _de
novo_ OTUs and compared the performance to existing tools.

@@

**TODO: ask Sarah Westcott to check the accuracy of this description**

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
OptiFit takes as input a list of reference OTUs and their sequences, a list of
query sequences to assign to the reference OTUs, the sequence pairs that are
within the distance threshold (e.g. 0.03), and the metric to assess clustering
quality (default: MCC).
Query sequences are randomly seeded in reference OTUs, then for each sequence
the algorithm calculates the quality metric based on whether the sequence stays
in its current OTU or moves to each of the other OTUs.
If two or more OTU assignments are of equal quality, a random number generator
is used to break the tie.
This process is repeated until the quality metric stabilizes, changing by no
more than 0.0001 by default, or until a maximum number of iterations is reached
(default: 100).
In closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: exactly what determines whether a seq can't be assigned?**), and
the results only contain OTUs that exist in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.

To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples. 
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.
For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
using both strategies.
To compare to existing software, we also clustered OTUs _de novo_ using
OptiClust and VSEARCH, and with VSEARCH in reference-based mode against the
Greengenes database.
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
We calculated the fraction of query sequences that were fit to existing OTUs in
closed reference mode as an additional measure of quality for this mode.

@@

To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by _de novo_ clustering the Greengenes
database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (**TODO: diff in MCC medians**).
OTU quality was remarkably stable across 100 different random seeds (**TODO:
variation or stdev?**).
In terms of runtime, closed reference OptiFit performed faster than OptiClust on
whole datasets by **Z%**.
In open reference mode, OptiClust performed **X to Y%** faster than OptiFit only
when the OptiFit reference fraction was 30% or less.
The split dataset strategy performed just as well as the database strategy in
open reference mode regardless of database used, and outperformed the database
strategy in closed reference mode by **W%**.

@@

The OptiClust and OptiFit algorithms provided by mothur produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to fit more sequence into OTUs than OptiFit in closed
reference mode.
While both mothur and VSEARCH use a distance or similarity threshold for
determining how to assign sequences to OTUs, VSEARCH is more permissive than
mothur.
Mothur's OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single sequence that seeded the OTU.
In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
fit to OTUs.
Users who require closed reference clustering to the Greengenes database may
prefer to use VSEARCH if they wish to maximize the fraction of sequences that
can be fit at the cost of OTU quality.
However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
reference clustering to produce OTUs of the highest possible quality.

@@

**TODO: big picture concluding paragraph.**
We have developed a new clustering algorithm that allows users to produce high
quality OTUs using already existing OTUs as a reference.

>>>>

- **TODO: current method for open/closed is vsearch against greengenes.**
- **TODO**: 2 categories of clustering: _de novo_ and reference based. separate paragraphs. describe opticlust first in de novo paragraph. 2nd paragraph: ref methods are good cause they're fast and don't use much ram. dependent on order of db. people use greengenes, which are rep seqs from 3% otus from full length.
- reader should know what opticlust is, closed & open ref clustering is, strengths & weakness are of each. then we solve these problems.
- **TODO:** note that greengenes is defunct now?!

To overcome the limitations of _de novo_ clustering while maintaining OTU
quality, we developed OptiFit, a reference-based clustering algorithm in the
mothur program which takes existing OTUs as the reference to fit new sequences
to. 
**TODO: more words here?** 
Here, we tested the OptiFit algorithm with the reference as a database or _de
novo_ OTUs and compared the performance to existing tools.
To evaluate the OptiFit algorithm and compare to existing methods, we used four
published datasets isolated from soil [@johnston_metagenomics_2016], marine
[@henson_artificial_2016], mouse gut [@schloss_stabilization_2012], and human
gut [@baxter_microbiota-based_2016] samples.

@@

- **TODO: ask Sarah Westcott to check the accuracy of this description**.
- **TODO: toy example like opticlust with figure**.

OptiFit leverages the method employed by OptiClust of iteratively assigning
sequences to OTUs to produce the highest quality OTUs possible, and extends this
method for reference-based clustering.
**TODO: more narrative, less documentation-like.**
**TODO: extension of opticlust except ref otus are fixed. just like opticlust only query seqs can move otus.**
OptiFit takes as input a list of reference OTUs and their sequences, a list of
query sequences to assign to the reference OTUs, the sequence pairs that are
within the distance threshold (e.g. 0.03), and the metric to assess clustering
quality (default: MCC).
Query sequences are randomly seeded in reference OTUs, then for each sequence
the algorithm calculates the quality metric based on whether the sequence stays
in its current OTU, moves to each of the other OTUs, or is discarded.
If two or more OTU assignments are of equal quality, a random number generator
is used to break the tie.
This process is repeated until the quality metric stabilizes, changing by no
more than 0.0001 by default, or until a maximum number of iterations is reached
(default: 100).
In closed-reference mode, any query sequences that cannot be assigned are thrown
out (**TODO: exactly what determines whether a seq can't be assigned?**), and
the results only contain OTUs that exist in the original reference.
In open-reference mode, unassigned query sequences are clustered _de novo_ using
OptiClust to generate additional OTUs. The final quality score is reported with
the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 1) fit sequences to
reference OTUs of an independent database, or 2) split the dataset into a
reference and query fraction, cluster the reference sequences _de novo_, then
fit the query sequences to the reference OTUs.

**TODO: make sure this info is baked into relevant sections, then remove.**
For each dataset repeated with 100 random seeds, we generated OTUs with OptiFit
using both strategies.
To compare to existing software, we also clustered OTUs _de novo_ using
OptiClust and VSEARCH, and with VSEARCH in reference-based mode against the
Greengenes database.

@@

We assigned blah ref dbs to 3% otus with opticlust. _de novo_ clustering the Greengenes
database, SILVA non-redundant database, or the Ribosomal Database Project (RDP)
[@desantis_greengenes_2006; @quast_silva_2013; @cole_ribosomal_2014].
All clustering was performed at a distance threshold of 0.03 and OTU quality was
evaluated using the MCC as described previously [@westcott_opticlust_2017].
To evaluate reference-based clustering with independent databases, we fit each
dataset to reference OTUs generated by
repeated 100 times.

TODO: use word "map" for what vsearch does, "fit" for what optifit does.

@@

OTU quality from the split dataset strategy with OptiFit was highly similar to
that from _de novo_ clustering the whole dataset with OptiClust regardless of
mode (**TODO: diff in MCC medians**).
OTU quality was remarkably stable across 100 different random seeds (**TODO:
coeff of variation?**).
In terms of runtime, closed reference OptiFit performed faster than OptiClust on
whole datasets by **Z%**.
In open reference mode, OptiClust performed **X to Y%** faster than OptiFit only
when the OptiFit reference fraction was 30% or less.
The split dataset strategy performed just as well as the database strategy in
open reference mode regardless of database used, and outperformed the database
strategy in closed reference mode by **W%**.

@@

The OptiClust and OptiFit algorithms provided by mothur produced higher quality
OTUs than VSEARCH in open reference, closed reference, or _de novo_ modes.
However, VSEARCH was able to fit more sequence into OTUs than OptiFit in closed
reference mode.
While both mothur and VSEARCH use a distance or similarity threshold for
determining how to assign sequences to OTUs, VSEARCH is more permissive than
mothur.
The OptiFit and OptiClust algorithms use all of the sequences to define an
OTU, requiring that all pairs of sequences (including reference and query
sequences) in an OTU are within the distance threshold without penalizing the
MCC.
In contrast, VSEARCH only requires each query sequence to be similar to the
single sequence that seeded the OTU.
In this way, VSEARCH sacrifices OTU quality in order to allow more sequences to
fit to OTUs.
Users who require closed reference clustering to the Greengenes database may
prefer to use VSEARCH if they wish to maximize the fraction of sequences that
can be fit at the cost of OTU quality.
However, mothur's OptiClust or OptiFit are recommended for _de novo_ or open
reference clustering to produce OTUs of the highest possible quality.

@@

**TODO: big picture concluding paragraph.**
We have developed a new clustering algorithm that allows users to produce high
quality OTUs using already existing OTUs as a reference.
**TODO: Point to courtney's paper metaphorically. wow what a cool application someone should do _wink wink_.**
